---
title: "Assignment_4"
author: "Nicole Szeluga"
date: "10/19/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Assignment 4: Data transformation with dplyr and visualization with
ggplot
================

<br>

## Instructions: Please read through this before you begin

  - This assignment is due by **10pm on Monday 10/19/20**.

  - For this assignment, please **reproduce this markdown file** using R
    markdown. This includes the followings:
    
      - **Reproduce this markdown template**. Pay attention to all the
        formating in this file, including bullet points, bolded
        characters, inserted code chunks, headings, text colors, blank
        lines, etc.
    
      - **Transform the data as instructed**. Try to use `tidyverse`
        functions even if you are more comfortable with base-R
        solutions. Show the **first 6 lines** of the transformed data in
        a table through RMarkdown **using the kable() function**, as
        shown in this markdown file.
    
      - **Reproduce the plots exactly as shown in this html file**. In
        two cases where the plot is not shown (Excercises 3.7 and 3.9),
        generate plots that you think can best answer the question.
    
      - Have all your code embedded within the R markdown file, and show
        **BOTH your code and plots** in the knitted markdown file.
    
      - When a verbal response is needed, answer by editing the part in
        the R markdown template where it says <span style="color:blue">
        “Write your response here” </span> .
    
      - Use R Markdown functionalities to **hide messages and warnings
        when needed**. (Suggestion: messages and warnings can often be
        informative and important, so please examine them carefully and
        only turn them off when you finish the exercise).

  - Please name your R markdown file `assignment_4.Rmd` and the knitted
    markdown file `assignment_4.md`. Please push both files to your
    class GitHub repository.

<br>

First, load all the required packages with the following code. Install
them if they are not installed yet.

```{r, results='hide'}
library(tidyverse)
library(knitr)
library(gapminder)
```

<br>

## Exercise 1. Theophylline experiment

This exercise uses the `Theoph` data frame (comes with your R
installation), which has 132 rows and 5 columns of data from an
experiment on the pharmacokinetics of the anti-asthmatic drug
theophylline. Twelve subjects were given oral doses of theophylline then
serum concentrations were measured at 11 time points over the next 25
hours. You can learn more about this dataset by running `?Theoph`

Have a look at the data structure

```{r}
kable(head(Theoph))
```

<br>

#### 1.1 Select columns that contain a lower case “t” in the `Theoph` dataset. Do not manually list all the columns to include.

```{r}
Theoph %>% 
  select(contains('t'), -Time) %>% 
  head(6)

```

<br>

#### 1.2 Rename the `Wt` column to `Weight` and `conc` column to `Concentration` in the `Theoph` dataset.

```{r}
Theoph %>% 
  rename(, Weight=Wt, Concentration=conc) %>% 
  head(6)
```

<br>

#### 1.3 Extract the `Dose` greater than 4.5 and `Time` greater than the mean `Time`.
```{r}
Theoph %>% 
  filter(Dose>4.5, Time> mean(Time)) %>% 
  head(6)
```

<br>

#### 1.4 Sort the `Theoph` dataset by `Wt` from smallest to largest and secondarily by Time from largest to smallest.
```{r}
Theoph %>% 
  arrange(-desc(Wt)) %>% 
  head(6)
```

<br>

#### 1.5 Create a new column called `Quantity` that equals to `Wt` x `Dose` in the `Theoph` dataset. This will tell you the absolute quantity of drug administered to the subject (in mg). Replace the `Dose` variable with `Quantity`.

```{r}
Theoph %>% 
  mutate(Quantity = Wt*Dose) %>% 
  relocate(Quantity, .after = Wt) %>% 
  head(6)
```

<br>

#### 1.6 Find the mean `conc` and sum of the `Dose` received by each test subject.

Show data for the 6 subjects with the smallest sum of `Dose` as below.
**Do not define new intermediate objects for this exercise; use pipes to
chain together functions. **

```{r}
group_by(Theoph, Subject) %>%
  summarize(mean(conc), sum(Dose)) %>%
  arrange(`sum(Dose)`) %>%
  head() %>%
  kable()
  

```

<br>

## Exercise 2. Trend in land value

This excercise uses a dataset that describes the trend of land value
(`Land.Value`), among other variables, in different states in the US
1975-2013. The states are grouped into four different regions, under the
variable `region`. This dataset was obtained from the Data Science
Services of Harvard
University.

```{r}
housing <- read_csv("https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/landdata_states.csv")
kable(head(housing)) 
```

<br>

#### 2.1 Washington DC was not assigned to a region in this dataset. According to the United States Census Bureau, however, DC is part of the South region. Here:

  - #### Change the region of DC to “South” (Hint: there are multiple ways to do this, but `mutate()` and `ifelse()` might be helpful)

  - #### Save this updated `region` variable together with `State`, `Date` and `Land.Value` into a new data frame.

  - #### Select the records from DC in this new data frame. How many records are there from DC? Show its first 6 lines.


Answer: 
```{r}
DC_housing <-housing %>%
  mutate(region = ifelse(State == "DC", "South", region)) %>%
  select(region,State,Land.Value,Date)

filter(DC_housing,State == "DC") %>% 
  head(6)

```


<br>

#### 2.2 Generate a dataframe that summarizes the mean land value of each region at each time point.
```{r}
mlv <- housing %>% 
  group_by(region, Date) %>% 
  drop_na() %>% 
  summarise( mean_land_value = mean(Land.Value))
mlv
```

<br>

#### 2.3 Using the dataframe above, plot the trend in mean land value of each region through time.

```{r}
ggplot(data = mlv) +
  geom_line(mapping = aes(x = Date, y = mean_land_value, color = region, ))
```


<br>

## Exercise 3. Life expectancy and GDP per capita 1952-2007

This exercise uses the `gapminder` dataset from the `gapminder` package.
It describes the life expectancy (`lifeExp`), GDP per capita
(`gdpPercap`), and population (`pop`) of 142 countries from 1952 to
2007. These countries can be grouped into 5 continents. As a reminder,
**reproduce the following plots exactly as shown**.

```{r}
kable(head(gapminder))
```


<br>

#### 3.1 Use a scatterplot to explore the relationship between per capita GDP (`gdpPercap`) and life expectancy (`lifeExp`).

```{r}
ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp ))
```


<br>

#### 3.2 Add a smoothing line to the previous plot.

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth(se=FALSE)
```


<br>

#### 3.3 Show each continent in a different color, and fit a separate smoothing line to each continent to identify differences in this relationship between continents. Turn off the confidence intervals.

```{r}
ggplot(data = gapminder,mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point()+
  geom_smooth(se=FALSE)
```


<br>

#### 3.4 Use faceting to solve the same problem. Show the confidence intervals in this plot

```{r}
ggplot(data = gapminder,mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point()+
  geom_smooth()+
  facet_wrap(~continent)

```


<br>

#### 3.5 Explore the trend in life expectancy through time in each continent. Color by continent.

```{r}
ggplot(data = gapminder, mapping = aes(x = year, y = lifeExp, color = continent)) +
  geom_line(mapping = aes(line = country), se=FALSE)+
  facet_wrap(~continent)
  
```


<br>

#### 3.6 From the previous plot, we see some abnormal trends in Asia and Africa, where the the life expectancy in some countries sharply dropped at certain time periods. Here, we look into what happened in Asia in more detail. First, create a new dataset by filtering only the Asian countries. Show the first 6 lines of this filtered dataset.

```{r}
Asia <- gapminder %>% 
  filter(continent == 'Asia') 
head(Asia,6) 

```

<br>

#### 3.7 Using the filtered dataset, identify the countries that had abnormal trends in life expectancy **by plotting**, and discuss historical events possibly explaining these trends. (Hint: facet by country)

```{r}
ggplot(data = Asia, mapping = aes(x = year, y = lifeExp))+
  geom_line(mapping = aes(line = country, color = country), se=FALSE) +
  facet_wrap(~country)
```

Answer: I notice Cambodia, Iraq, ans Chine have odd trend lines that face a sharp change. This could be due to war or in china's case, enaction of the 1 child law. I suck at history. 


<br>

#### 3.8 Explore the trend in per capita GDP through time in each continent.

```{r}
ggplot(data= gapminder, mapping = aes( x = year, y = gdpPercap, color = continent))+
  geom_line(mapping = aes(line = country), se = FALSE)+
  facet_wrap(~continent)
```


<br>

#### 3.9 There is one Asian country that had a very sharp decline in per capita GDP. With the previously filtered dataset, **use a plot** to identify this country and speculate on the historical event underlying this pattern.

<br>
```{r}
ggplot(data= Asia, mapping = aes( x= year, y=gdpPercap, color = country))+
  geom_line(mapping = aes(line = country), se = FALSE)+
  facet_wrap(~country)

```
Answer: Kuwait had a very sharp decline in gdp per capita. This happened between 1970 and 1975. I think there was a change in political power during this time. 

<br>